### 2023-03-02 calculus.tex
Describe reified hazelnut action semantics

The marking rule mirrors \textsc{USPlus} closely. Since an expected type for each operand is known,
we shift the responsibility for any type errors to them. Hence, we recursively mark each operand in
analytic mode and rebuild the marked addition expression. The marked typing rule then mirrors
\textsc{USPlus} exactly.
%
\begin{mathpar}
  \judgment{
    \ctxAnaFixedInto{\ctx}{\EMV_1}{\ECMV_1}{\TNum} \\\\
    \ctxAnaFixedInto{\ctx}{\EMV_2}{\ECMV_2}{\TNum}
  }{
    \ctxSynFixedInto{\ctx}{\EPlus{\EMV_1}{\EMV_2}}{\ECPlus{\ECMV_1}{\ECMV_2}}{\TNum}
  }{MKSPlus}

  \judgment{
    \ctxAnaTypeM{\ctx}{\ECMV_1}{\TNum} \\\\
    \ctxAnaTypeM{\ctx}{\ECMV_2}{\TNum}
  }{
    \ctxSynTypeM{\ctx}{\ECPlus{\ECMV_1}{\ECMV_2}}{\TNum}
  }{MSPlus}
\end{mathpar}

@@

Again, to satisfy totality, we must consider the case when $\TMV$ is not a matched arrow type, such
as in the example of \cref{fig:calculus-examples-app-non-lambda}. There is no expected type for the
argument, so we perform analytic marking on $\EMV_2$ against the unknown type. In this case, it is
not quite right to mark $\EMV_1$ with the inconsistent type mark; rather than any single type, it is
a member of the family of function types that is expected. For such a ``constrained'' synthetic
mode, we use a specialized mark $\ECSynNonMatchedArrow{\ECMV}$, which indicates that $\ECMV$ was
expected to be a function---but was not. The output type is unknown, so the entire expression
synthesizes the unknown type.
%
\begin{mathpar}
  \judgment{
    \ctxSynFixedInto{\ctx}{\EMV_1}{\ECMV_1}{\TMV} \\
    \notMatchedArrow{\TMV} \\\\
    \ctxAnaFixedInto{\ctx}{\EMV_2}{\ECMV_2}{\TUnknown}
  }{
    \ctxSynFixedInto{\ctx}{\EAp{\EMV_1}{\EMV_2}}{\ECApSynNonMatchedArrow{\ECMV_1}{\ECMV_2}}{\TUnknown}
  }{MKSAp2}

  \judgment{
    \ctxSynTypeM{\ctx}{\ECMV_1}{\TMV} \\
    \notMatchedArrow{\TMV} \\\\
    \ctxAnaTypeM{\ctx}{\ECMV_2}{\TUnknown}
  }{
    \ctxSynTypeM{\ctx}{\ECApSynNonMatchedArrow{\ECMV}{\ECMV}}{\TUnknown}
  }{MSAp2}
\end{mathpar}

@@

  \judgment{
    \ctxSynTypeM{\ctx}{\ECMV}{\TMV} \\
    \matchedProd{\TMV}{\TMV_1}{\TMV_2}
  }{
    \ctxSynTypeM{\ctx}{\ECProjL{\ECMV}}{\TMV_1}
  }{MSProjL1}

  \judgment{
    \ctxSynTypeM{\ctx}{\ECMV}{\TMV} \\
    \matchedProd{\TMV}{\TMV_1}{\TMV_2}
  }{
    \ctxSynTypeM{\ctx}{\ECProjR{\ECMV}}{\TMV_2}
  }{MSProjR1}

  \judgment{
    \matchedProd{\TMV}{\TMV_1}{\TMV_2} \\
    \ctxAnaTypeM{\ctx}{\ECMV_1}{\TMV_1} \\
    \ctxAnaTypeM{\ctx}{\ECMV_2}{\TMV_2}
  }{
    \ctxAnaTypeM{\ctx}{\ECPair{\ECMV_1}{\ECMV_2}}{\TMV}
  }{MAPair1}

@@

Hazelnut, however, is unable to cope with actions that require non-local hole fixes. % TODO example of this
Such actions remain undefined, i.e. unperformable.

By applying the marked lambda calculus to Hazelnut, the action calculus may be \emph{wholly decoupled}
from the typing semantics. Under this new system, the action calculus is concerned only with the
manipulation of syntax, and the marking procedure yields statically meaningful terms with error
holes inserted at the appropriate positions. Furthermore, marking solves the problem of non-local
hole fixes; actions may be performed anywhere (as long as they are of the correct sort), and marking
totality ensures that they give meaningful terms.
% TODO this is not strictly true, only for 'wrapping' actions

>>>>

The marking rule mirrors \textsc{USPlus} closely. Since an expected type for each operand is known,
we shift the responsibility for any type errors to them. Hence, we recursively mark each operand in
analytic mode and rebuild the marked addition expression. The marked typing rule then mirrors
\textsc{USPlus} exactly.
%
\begin{mathpar}
  \judgment{
    \ctxAnaFixedInto{\ctx}{\EMV_1}{\ECMV_1}{\TNum} \\
    \ctxAnaFixedInto{\ctx}{\EMV_2}{\ECMV_2}{\TNum}
  }{
    \ctxSynFixedInto{\ctx}{\EPlus{\EMV_1}{\EMV_2}}{\ECPlus{\ECMV_1}{\ECMV_2}}{\TNum}
  }{MKSPlus}

  \judgment{
    \ctxAnaTypeM{\ctx}{\ECMV_1}{\TNum} \\
    \ctxAnaTypeM{\ctx}{\ECMV_2}{\TNum}
  }{
    \ctxSynTypeM{\ctx}{\ECPlus{\ECMV_1}{\ECMV_2}}{\TNum}
  }{MSPlus}
\end{mathpar}

@@

Again, to satisfy totality, we must consider the case when $\TMV$ is not a matched arrow type, such
as in the example of \cref{fig:calculus-examples-app-non-lambda}. There is no expected type for the
argument, so we perform analytic marking on $\EMV_2$ against the unknown type. In this case, it is
not quite right to mark $\EMV_1$ with the inconsistent type mark; rather than any single type, it is
a member of the family of function types that is expected. For such a ``constrained'' synthetic
mode, we use a specialized mark $\ECSynNonMatchedArrow{\ECMV}$, which indicates that $\ECMV$ was
expected to be a function---but was not. The output type is unknown, so the entire expression
synthesizes the unknown type.
%
\begin{mathpar}
  \judgment{
    \ctxSynFixedInto{\ctx}{\EMV_1}{\ECMV_1}{\TMV} \\
    \notMatchedArrow{\TMV} \\
    \ctxAnaFixedInto{\ctx}{\EMV_2}{\ECMV_2}{\TUnknown}
  }{
    \ctxSynFixedInto{\ctx}{\EAp{\EMV_1}{\EMV_2}}{\ECApSynNonMatchedArrow{\ECMV_1}{\ECMV_2}}{\TUnknown}
  }{MKSAp2}

  \judgment{
    \ctxSynTypeM{\ctx}{\ECMV_1}{\TMV} \\
    \notMatchedArrow{\TMV} \\
    \ctxAnaTypeM{\ctx}{\ECMV_2}{\TUnknown}
  }{
    \ctxSynTypeM{\ctx}{\ECApSynNonMatchedArrow{\ECMV}{\ECMV}}{\TUnknown}
  }{MSAp2}
\end{mathpar}

@@

  \judgment{
    \ctxSynTypeM{\ctx}{\ECMV}{\TMV} \\\\
    \matchedProd{\TMV}{\TMV_1}{\TMV_2}
  }{
    \ctxSynTypeM{\ctx}{\ECProjL{\ECMV}}{\TMV_1}
  }{MSProjL1}

  \judgment{
    \ctxSynTypeM{\ctx}{\ECMV}{\TMV} \\\\
    \matchedProd{\TMV}{\TMV_1}{\TMV_2}
  }{
    \ctxSynTypeM{\ctx}{\ECProjR{\ECMV}}{\TMV_2}
  }{MSProjR1}

  \judgment{
    \matchedProd{\TMV}{\TMV_1}{\TMV_2} \\
    \ctxAnaTypeM{\ctx}{\ECMV_1}{\TMV_1} \\\\
    \ctxAnaTypeM{\ctx}{\ECMV_2}{\TMV_2}
  }{
    \ctxAnaTypeM{\ctx}{\ECPair{\ECMV_1}{\ECMV_2}}{\TMV}
  }{MAPair1}

@@

Hazelnut, however, is unable to cope with actions that require \emph{non-local} hole fixes. For
example, though the system can wrap an arbitrary expression in an arithmetic operation, it cannot
wrap an arbitrary expression $\EMV$ in a lambda abstraction, i.e. construct
$\ELam{x}{\TUnknown}{\EMV}$ directly from $\EMV$. The binding of $x$ with the unknown type may
shadow a previous declaration and require the removal of error holes anywhere in body, but Hazelnut
may only perform such fixes for immediately local holes. Such actions remain undefined, i.e.
unperformable.

By applying the marked lambda calculus to Hazelnut, the action calculus may be \emph{wholly decoupled}
from the typing semantics. Under this new system, the action calculus is concerned only with the
manipulation of syntax, and the marking procedure yields statically meaningful terms with error
marks, now error holes, inserted at the appropriate positions as a kind of editor decoration. This
solves the problem outlined above; since marking operates on the entire program after each action,
arbitrary constructions are permitted.

Instead of a wholly untyped action calculus, one may re-adapt the existing Hazelnut calculus with
scope-based marking. Action operate over \emph{marked expressions}, and error marks operate as error
holes. When wrapping $\ECMV$ into $\ECLam{x}{\TUnknown}{\ECMV}$, we may re-mark the body under the
context with $x$ binding type $\TUnknown$:
\[%
  \judgment{
    \ctxSynFixedInto{\extendCtx{\ctx}{x}{\TUnknown}}{\erase{\ECMV}}{\ECMV'}{\TMV'}
  }{
    \ASEConLam{\ctx}{\ECMV}{\TMV}{\ECMV'}{\TMV'}{x}
  }{ASEConLam}
\]%
This system optimizes over the untyped version by performing marking only when necessary. In
particular, whenever binders are introduced around an expression, marking redraws error holes. The
supplemental material contains a complete formal description of both variants, and the untyped
action semantics and related metatheory are mechanized in Agda.
% TODO note that we have not verified the metatheory for the typed system
